Supplementary material for: non-coding RNAs of bird genomes

### 2013-11-05 paper/manuscript.tex
Merged comments from Peter, Sarah, Sam & Mario.

\author{
Paul P. Gardner\correspondingauthor$^{1,2}$
\email{Paul P. Gardner\correspondingauthor - paul.gardner@canterbury.ac.nz},
Jana Hertel$^5$
\email{Jana Hertel\correspondingauthor - jana@bioinf.uni-leipzig.de},
Sarah W. Burge$^3$
\email{sb30@sanger.ac.uk},
Maria Ninova$^4$
\email{Maria.Ninova@postgrad.manchester.ac.uk},
Stephanie Kehr$^5$
\email{steffi@bierdepot.bioinf.uni-leipzig.de},
Mario Fasold$^5$
\email{mario@bierdepot.bioinf.uni-leipzig.de},
Tammy E. Steeves$^1$
\email{tammy.steeves@canterbury.ac.nz},
Sam Griffiths-Jones$^4$
\email{sam.griffiths-jones@manchester.ac.uk}
and
Peter Stadler\correspondingauthor$^5$
\email{Peter Stadler\correspondingauthor - studla@bioinf.uni-leipzig.de}
}
\address{
\iid(1) School of Biological Sciences, University of Canterbury, Private Bag 4800, Christchurch, New Zealand.
\iid(2) Biomolecular Interaction Centre, University of Canterbury, Private Bag 4800, Christchurch, New Zealand.
\iid(3) European Molecular Biology Laboratory, European Bioinformatics Institute, Hinxton, Cambridge, CB10 1SD, UK.
\iid(4) Faculty of Life Sciences, University of Manchester, Manchester, United Kingdom.
\iid(5) Bioinformatics Group, Department of Computer Science; and Interdisciplinary Center for Bioinformatics, University of Leipzig, H{\"a}rtelstrasse 16-18, D-04107 Leipzig, Germany
}

@@

\begin{abstract}
Here we present the results of a large-scale bioinformatic annotation
of non-coding RNAs in 48 avian genomes. Our approach uses
probabilistic models of hand-curated families from the Rfam database
to infer conserved RNA families within each avian genome. We
supplement these annotations with predictions from the tRNA annotation
tool, tRNAscan-SE and microRNAs from miRBase.  Although we show a
significant number of lncRNAs are surprisingly well conserved between
birds and mammals including several intriguing cases where the
reported mammalian lncRNA function is not conserved in birds.  In
contrast, we show extensive conservation of classical ncRNAs (e.g.,
tRNAs and rRNAs) and more recently discovered ncRNAs (e.g., snoRNAs
and miRNAs) in birds, we also demonstrate apparent ``losses'' in
several RNA families. These include the divergence of some classical
ncRNAs and the loss of several snoRNAs and microRNAs.  These combined
results illustrate the utility of applying homology based methods for
annotating vertebrate genomes and illustrate many complex evolutionary
patterns within the avian ncRNA cohort.
\end{abstract}

@@

Non-coding RNAs (ncRNAs) are an important class of genes. However,
they pose serious research challenges, particularly for the field of
genomics. For example, they lack the strong statistical signals
associated with protein coding genes, e.g. open reading frames, G+C
content and codon-usage biases \cite{Rivas:2000}.

Non-coding RNAs are classified into hierarchical groupings by the Rfam
database. The basic units are ``families'' which are groups of
homologous, alignable sequences; ``clans'' which are groups of
un-alignable (or functionally distinct), homologous families; and
``classes'' which are groups of clans and families with related
biological functions e.g. spliceosomal RNAs, miRNAs and snoRNAs
\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}.
The major RNA families include the classical, highly conserved RNAs,
sometimes called ``molecular fossils'', such as the transfer RNAs,
ribosomal RNAs, RNA components of RNase P and the signal recognition
particle \cite{Jeffares:1998}. Other classes appear to have have
evolved more recently, e.g. the small nucleolar RNAs (snoRNAs),
microRNAs (miRNAs) and the long non-coding RNAs (lncRNAs)
\cite{Hoeppner:2012}. The publication of 48 avian genomes, including
the previously published chicken
\cite{International_Chicken_Genome_Sequencing_Consortium:2004}, zebra
finch \cite{Warren:2010} and turkey \cite{Dalloul:2010} with the newly
published 45 avian genomes \cite{}, provides exciting opportunities to
to explore ncRNA conservation in unprecedented detail.

Despite promising techniques such as RNA-seq \cite{Croucher:2010},
homology based methods namely covariance models (CMs) remain state of the
art for ncRNA analyses \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.
The CM based approach for annotating ncRNAs in genomes requires
reliable alignments and consensus secondary structures of
representative sequences of RNA families. These are used to train
probabilistic models for each family. These models can be used to
generate sequences with similar properties, score the likelihood that
a sequence is was generated by the same evolutionary processes as the
training sequences and to build alignments based upon sequence and
structural information \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.
The tRNAscan-SE software package uses CMs to accurately predict
transfer RNAs \cite{Lowe:1997,Chan:2009}. The Rfam database contains
thousands of curated alignments and consensus structures for diverse
classes of ncRNAs
\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}.
Independent benchmarks of bioinformatic annotation tools have shown
that the CM approaches dramatically out-perform alternative methods
\cite{Freyhult:2007}.

The CM based approach works well for almost all classes of ncRNA, but
the long non-coding RNAs (lncRNAs) are a particular challenge
\cite{Guttman:2009}. Recent technological advances have led to dramatic
speed and memory-usage enhancements for CM analyses
\cite{Eddy:2002,Nawrocki:2007,Nawrocki:2009,Eddy:2011}. However, CMs
cannot model the exon-intron structures of spliced lncRNAs, nor can they
deal simply with the repeats that many lncRNAs host. Consequently in
the latest release of Rfam the lncRNA families that were added were
composed of local conserved and structured elements within lncRNAs,
analogous to the ``domains'' housed within protein sequences
\cite{Burge:2013}. The functions determined to date for lncRNAs
range from regulating chromatin status to chromosomal inactivation
\cite{Rinn:2007,Chow:2005}. Yet functional characterisation of these
genes is a lengthy and expensive process \cite{Guttman:2009}.

@@

\subsection*{Unusually, well conserved RNAs}

The bulk of the ``unusually well conserved RNAs'' belong to the long
non-coding RNA (lncRNA) group.  The lncRNAs are a diverse group of
RNAs that have been implicated in a multitude of functional processes
\cite{Rinn:2007,Chow:2005,Guttman:2009}. These RNAs have largely been
characterised in mammalian species, particularly human and
mouse. Consequently, we generally do not expect these to be conserved
outside of mammalia. Notable examples include Xist \cite{Duret:2006}
and H19 \cite{Smits:2008}.  There is emerging evidence for the
conservation of “mammalian” lncRNAs in other vertebrates
\cite{Chodroff:2010,Ulitsky:2011}), however, like most lncRNAs, the
function of these lncRNAs remains largely unknown. Here, we show the
conservation of several well-characterised lncRNAs of known function
in humans.

@@

When analysing the RNA-domain annotations it is striking that many of
the lncRNAs with multiple RNA-domains are consistently preserved in
the birds. The annotations of these domains lie in the same genomic
region, in the same order as in the mammalian homologs. Thus
supporting a high degree of evolutionary conservation for the entire
lncRNA. In particular the HOXA11-AS1, PART1, PCA3, RMST, Six3os1, SOX2OT and
ST7-OT3 lncRNAs have multiple, well conserved RNA-domains (See
Figure~\ref{fig:5}).

@@

The HOX cluster lncRNAs, HOTAIRM1, HOTTIP and HOXA11-AS1, are
remarkably well conserved. In the human genome the corresponding
RNA-domains are located at chr7:27135743-27245922 (hg19) and are
modelled by 15 RNA-domains in total, the 
lncRNAs are ordered HOTAIRM1 (5 RNA-domains) $\rightarrow$ HOXA11-AS1
(6 RNA-domains) $\rightarrow$ HOTTIP (4 RNA-domains). In the alligator
and turtle genomes five and six of the HOX RNA-domains are conserved,
each supporting an ancestral gene order of HOTAIRM1 $\rightarrow$
HOXA11-AS1 $\rightarrow$ HOTTIP. In the birds, where two or more of
the HOX cluster lncRNA RNA-domains were predicted on the same
scaffold, this gene order was also preserved.
%egrep 'HOXA11-AS1|HOTAIR|HOTTIP' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'
%egrep 'PCA3' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'
%egrep 'RMST' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'

@@

The simplest case are the selenocystein tRNAs. Here, tRNAscan is tuned for
specificity and thus misses several occurrances that are easily found by
\texttt{blastn} with $E\le 10^{-30}$. In some cases the sequences appear
degraded at the ends, which may be explained e.g.\ by low sequence quality
at the very ends of contigs or scaffolds. A \texttt{blastn} search also
readily retrieves additional RNAse P and RNAse MRP RNAs, capturing only the
best conserved regions. In many cases these additional candidates are
incomplete or contain undetermined sequence, explaining why they are missed
by the CMs. Overall, we identify tRNA-Sec in most and RNAse P and MRP RNAs
in the majority of the genomes. An additional candidate could also be
retrieved for telomerase RNA. Telomerase is well known to exhibit very poor
sequence conservation and rapid variations in size that make it notoriously
hard to identify by homology search \cite{Xie:08a}. The poor return thus
does not come as a surprise. Since \texttt{blastn} searches remained
unsuccessfull we constructed a sauropsid-specific CM for the vault RNA. In
addition to the hits identified by the Rfam model we obtained three
additional homologs. Vault RNAs, with a size of about 100 nt, exhibit
conserved sequence patterns only at their ends, with essentially
unconstrained sequence in the central part. Their identification is also a
well-know difficult problem for homology search
\cite{Stadler:09b}.

Our ability to find additional homologs for several RNA families that fill
gaps in the abundance matrices (Figure~\ref{fig:7})
strongly suggests that that conspicuous absences, in particular of LUCA and
LECA RNAs, are caused by incomplete data in the current assemblies and
sequence divergence rather then true losses.

@@

The microRNA family let-7 is the most abundant microRNA family with 14
paralogs in human. These genes also localize in 7 genomic clusters,
together with mir-100 and mir-125 miRNA families (see previous study on the
evolution of the let-7 miRNA cluster in \cite{Hertel:2012}). In Sauropsids
we observed that cluster A - that is strongly conserved in vertebrates has
been completely lost in the avian lineage. Another obvious loss in birds is
cluster F, containing two let-7 microRNA paralogs. Cluster H, on the other
hand has been retained in all oviparous animals and completely lost later,
after the split of Theria. See Supplemental Figure~7 for details.

@@

In this work we have summarised the main results after carefully
annotating non-coding RNAs in genome sequences using homology-based
methods. The homology-based tools have distinct advantages over
experimental-based approaches as not all RNAs are expressed in any
particular tissue-type or developmental-stage, in fact some RNAs have
extremely specific expression profiles \cite{Johnston:2003}.

 We have identified previously unrecognised conservation of
ncRNAs in avian genomes as well as some surprising ``losses'' of
otherwise well conserved ncRNAs. We suspect many of these losses are
due to a combination of limitations in the homology search tools that
we use for annotation and the ability of ncRNAs to tolerate large
amounts of sequence variation while remaining functional, rather than
\emph{bona fide} gene loss. In some cases these losses could be due to
missing data from the genome assemblies, but this unlikely to be the
case for multiple independant assemblies.

@@

Bird genomes were searched using the cmsearch program from INFERNAL
1.1 and the covariance models from the Rfam database
v11.0 \cite{Gardner:2011a,Burge:2013}. All matches above the curated GA
threshold were included. Subsequently, all hits with an E-value
greater than 0.0005 were discarded, so only matches which passed the
model-specific GA threshold and had an E-value smaller than 0.0005
were retained.

@@

SnoRNA annotations overlapping miRNAs annotations where manually
inspected \ppg{and?...}.

The snoRNA and miRNA annotations were validated using RNA-seq data
(\ppg{SOURCED FROM?}). Since not all RNAs are expressed at any given
time we expect to verify a fraction of RNAs annotated using expression
data. Never-the-less 242 of 691 (35\%) miRNAs validated as well as 328
of 376 (87\%) snoRNAs of our homology-based annotations were validated
using this data.

@@

Erich Jarvis, Guojie Zhang and Tom Gilbert. Magnus Alm Rosenblad. 
%%%%%%%%%%%%%%%%%%%%%%
%% The Bibliography %%
%%
%% ---------------------------------
%% BioMedCentral bibtex .BST file will be used to
%% create a .BBL file which includes the BMC XML.
%% Note that the displayed Bibliography will not be
%% exactly as specified in the online Instructions for Authors

@@

\begin{figure}[ht]
  \includegraphics[width=0.75\textwidth]{figures/More-cand.pdf}
  \caption[]{...}\label{fig:7}
\end{figure}

>>>>

\author{
Paul P.\ Gardner\correspondingauthor$^{1,2}$
\email{Paul P.\ Gardner\correspondingauthor - paul.gardner@canterbury.ac.nz},
Jana Hertel$^5$
\email{Jana Hertel\correspondingauthor - jana@bioinf.uni-leipzig.de},
Sarah W.\ Burge$^3$
\email{swb@ebi.ac.uk},
Maria Ninova$^4$
\email{Maria.Ninova@postgrad.manchester.ac.uk},
Stephanie Kehr$^5$
\email{steffi@bierdepot.bioinf.uni-leipzig.de},
Mario Fasold$^5$
\email{mario@bierdepot.bioinf.uni-leipzig.de},
Tammy E.\ Steeves$^1$
\email{tammy.steeves@canterbury.ac.nz},
Sam Griffiths-Jones$^4$
\email{sam.griffiths-jones@manchester.ac.uk}
and
Peter F.\ Stadler\correspondingauthor$^5$
\email{Peter Stadler\correspondingauthor - studla@bioinf.uni-leipzig.de}
}
\address{
\iid(1) School of Biological Sciences, University of Canterbury, Private Bag 4800, Christchurch, New Zealand.
\iid(2) Biomolecular Interaction Centre, University of Canterbury, Private Bag 4800, Christchurch, New Zealand.
\iid(3) European Molecular Biology Laboratory, European Bioinformatics Institute, Hinxton, Cambridge, CB10 1SD, UK.
\iid(4) Faculty of Life Sciences, University of Manchester, Manchester, United Kingdom.
\iid(5) Bioinformatics Group, Department of Computer Science; and Interdisciplinary Center for Bioinformatics, University of Leipzig, H{\"a}rtelstrasse 16-18, D-04107 Leipzig, Germany
}

@@

\begin{abstract}
Here we present the results of a large-scale bioinformatic annotation
of non-coding RNAs in 48 avian genomes. Our approach uses
probabilistic models of hand-curated families from the Rfam database
to infer conserved RNA families within each avian genome. We
supplement these annotations with predictions from the tRNA annotation
tool, tRNAscan-SE and microRNAs from miRBase.  Although we show a
significant number of lncRNAs are surprisingly well conserved between
birds and mammals including several intriguing cases where the
reported mammalian lncRNA function is not conserved in birds.  In
contrast, we show extensive conservation of classical ncRNAs (e.g.,
tRNAs) and more recently discovered ncRNAs (e.g., snoRNAs
and miRNAs) in birds, we also demonstrate apparent ``losses'' in
several RNA families. These include the divergence of some classical
ncRNAs and the loss of several snoRNAs and microRNAs.  These combined
results illustrate the utility of applying homology based methods for
annotating vertebrate genomes and illustrate many complex evolutionary
patterns within the avian ncRNA cohort.
\end{abstract}

@@

Non-coding RNAs (ncRNAs) are an important class of genes, responsible
for the regulation of many key cellular functions. The major RNA
families include the classical, highly conserved RNAs, sometimes
called ``molecular fossils'', such as the transfer RNAs, ribosomal
RNAs, RNA components of RNase P and the signal recognition particle
\cite{Jeffares:1998}. Other classes appear to have have evolved more
recently, e.g. the small nucleolar RNAs (snoRNAs), microRNAs (miRNAs)
and the long non-coding RNAs (lncRNAs) \cite{Hoeppner:2012}.

The ncRNAs pose serious research challenges, particularly for the
field of genomics. For example, they lack the strong statistical
signals associated with protein coding genes, e.g. open reading
frames, G+C content and codon-usage biases \cite{Rivas:2000}. Despite
promising techniques such as RNA-seq \cite{Croucher:2010}, homology
based methods, namely covariance models (CMs) remain state of the art
for ncRNA analyses \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.
The CM based approach for annotating ncRNAs in genomes requires
reliable alignments and consensus secondary structures of
representative sequences of RNA families. These are used to train
probabilistic models for each family. These models can be used to
generate sequences with similar properties, score the likelihood that
a sequence is generated by the same evolutionary processes as the
training sequences and to build alignments based upon sequence and
structural information \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.
The tRNAscan-SE software package uses CMs to accurately predict
transfer RNAs \cite{Lowe:1997,Chan:2009}.  The Rfam database contains
thousands of curated alignments and consensus structures for diverse
classes of ncRNAs
\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}.
Independent benchmarks of bioinformatic annotation tools have shown
that the CM approaches dramatically out-perform alternative methods
\cite{Freyhult:2007}, although its sensitivity is limited for the most
rapidly evolving families such as vault RNAs or telomerase RNA
\cite{Menzel:09a}.

The CM based approach works well for almost all classes of ncRNA, but
the long non-coding RNAs (lncRNAs) are a particular challenge
\cite{Guttman:2009}. Recent technological advances have led to
dramatic speed and memory-usage enhancements for CM analyses
\cite{Eddy:2002,Nawrocki:2007,Nawrocki:2009,Eddy:2011}. However, CMs
cannot model the exon-intron structures of spliced lncRNAs, nor can
they deal simply with the repeats that many lncRNAs host. Consequently
in the latest release of Rfam the lncRNA families that were added were
composed of local conserved (and possibly structured elements) within
lncRNAs, analogous to the ``domains'' housed within protein sequences
\cite{Burge:2013}. The functions determined to date for lncRNAs range
from regulating chromatin status to chromosomal inactivation
\cite{Rinn:2007,Chow:2005}. Yet functional characterisation of these
genes is a lengthy and expensive process \cite{Guttman:2009}.

The publication of 48 avian genomes, including the previously
published chicken
\cite{International_Chicken_Genome_Sequencing_Consortium:2004}, zebra
finch \cite{Warren:2010} and turkey \cite{Dalloul:2010} with the newly
published 45 avian genomes \cite{birds:14,birds:14a}, provides
exciting opportunities to to explore ncRNA conservation in
unprecedented detail.

@@

\subsection*{Unusually well conserved RNAs}

The bulk of the ``unusually well conserved RNAs'' belong to the long
non-coding RNA (lncRNA) group.  The lncRNAs are a diverse group of
RNAs that have been implicated in a multitude of functional processes
\cite{Rinn:2007,Chow:2005,Guttman:2009}. These RNAs have largely been
characterised in mammalian species, particularly human and
mouse. Consequently, we generally do not expect these to be conserved
outside of mammalia. Notable examples include Xist \cite{Duret:2006}
and H19 \cite{Smits:2008}.  There is emerging evidence for the
conservation of ``mammalian'' lncRNAs in other vertebrates
\cite{Chodroff:2010,Ulitsky:2011}), however, like most lncRNAs, the
function of these lncRNAs remains largely unknown. Here, we show the
conservation of several well-characterised lncRNAs of known function
in humans.

@@

When analysing the RNA-domain annotations it is striking that many of
the lncRNAs with multiple RNA-domains are consistently preserved in
the birds. The annotations of these domains lie in the same genomic
region, in the same order as in the mammalian homologs. Thus they
support a high degree of evolutionary conservation for the entire
lncRNA. In particular the HOXA11-AS1, PART1, PCA3, RMST, Six3os1, SOX2OT and
ST7-OT3 lncRNAs have multiple, well conserved RNA-domains (See
Figure~\ref{fig:5}).

@@

The HOX cluster lncRNAs HOTAIRM1 (5 RNA-domains), HOXA11-AS1 (6
RNA-domains), and HOTTIP (4 RNA domains) are remarkably well conserved. In
the human genome they are located in the HOXA cluster (hg coordinates
chr7:27135743-27245922), one of the most highly conserved regions in
vertebrate genomes \cite{PascualAnaya:13}, in antisense orientation between
HoxA1 and HoxA2, between HoxA11 and HoxA13, and upstream of HoxA13,
respectively. Conservation and expression of HOTAIRM1 and HOXA11-AS1 within
the HOXA cluster has been studied in some detail in marsupials
\cite{Yu:12}.  Of the 15 RNA-domains five and six representing all three
lncRNAs were recovered in the alligator and turtle genomes. All of them
appear in the correct order at the expected, syntenically conserved
positions within the HOXA cluster.  In the birds, where two or more of the
HOX cluster lncRNA RNA-domains were predicted on the same scaffold, this
gene order and location within HOX was also preserved.

%egrep 'HOXA11-AS1|HOTAIR|HOTTIP' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'
%egrep 'PCA3' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'
%egrep 'RMST' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'

@@

The simplest case are the selenocystein tRNAs. Here, tRNAscan is tuned
for specificity and thus misses several occurrances that are easily
found by \texttt{blastn} with $E\le 10^{-30}$. In some cases the
sequences appear degraded at the ends, which may be explained e.g.\ by
low sequence quality at the very ends of contigs or scaffolds. A
\texttt{blastn} search also readily retrieves additional RNAse P and
RNAse MRP RNAs, capturing only the best conserved regions. In many
cases these additional candidates are incomplete or contain
undetermined sequence, explaining why they are missed by the
CMs. Overall, we identify tRNA-Sec in most and RNAse P and MRP RNAs in
the majority of the genomes. An additional candidate could also be
retrieved for telomerase RNA. Telomerase is well known to exhibit very
poor sequence conservation and rapid variations in size that make it
notoriously hard to identify by homology search \cite{Xie:08a}. The
poor return thus does not come as a surprise. Since \texttt{blastn}
searches remained unsuccessfull we constructed a sauropsid-specific CM
for the vault RNA. In addition to the hits identified by the Rfam
model we obtained three additional homologs. Vault RNAs, with a size
of about 100 nt, exhibit conserved sequence patterns only at their
ends, with essentially unconstrained sequence in the central
part. Their identification is one of the well-know difficult problems
for homology search \cite{Stadler:09b}.

Our ability to find additional homologs for several RNA families that fill
gaps in the abundance matrices (Figure~\ref{fig:7})
strongly suggests that conspicuous absences, in particular of LUCA and
LECA RNAs, are caused by incomplete data in the current assemblies and
sequence divergence rather then true losses.

@@

The microRNA family let-7 is the most diverse microRNA family with 14
paralogs in human. These genes also localize in 7 genomic clusters,
together with mir-100 and mir-125 miRNA families (see previous study on the
evolution of the let-7 miRNA cluster in \cite{Hertel:2012}). In Sauropsids
we observed that cluster A - which is strongly conserved in vertebrates has
been completely lost in the avian lineage. Another obvious loss in birds is
cluster F, containing two let-7 microRNA paralogs. Cluster H, on the other
hand has been retained in all oviparous animals and completely lost later,
after the split of Theria. See Supplemental Figure~7 for details.

@@

In this work we have provided a comprehensive annotation of non-coding
RNAs in genome sequences using homology-based methods. The
homology-based tools have distinct advantages over experimental-based
approaches as not all RNAs are expressed in any particular tissue-type
or developmental-stage, in fact some RNAs have extremely specific
expression profiles \cite{Johnston:2003}.  We have identified
previously unrecognised conservation of ncRNAs in avian genomes as
well as some surprising ``losses'' of otherwise well conserved
ncRNAs. We suspect many of these losses are due to a combination of
limitations in the homology search tools that we use for annotation
and the ability of ncRNAs to tolerate large amounts of sequence
variation while remaining functional, rather than \emph{bona fide}
gene loss. In some cases these losses could be due to missing data
from the genome assemblies, but this unlikely to be the case for
multiple independant assemblies.

@@

Bird genomes were searched using the cmsearch program from INFERNAL
1.1 and the covariance models from the Rfam database
v11.0 \cite{Gardner:2011a,Burge:2013}. All matches above the curated GA
threshold were included. Subsequently, all hits with an E-value
greater than 0.0005 were discarded, so only matches which passed the
model-specific GA threshold and had an E-value smaller than 0.0005
were retained. The Rfam database classifies non-coding RNAs into
hierarchial groupings. The basic units are ``families'' which are
groups of homologous, alignable sequences; ``clans'' which are groups
of un-alignable (or functionally distinct), homologous families; and
``classes'' which are groups of clans and families with related
biological functions e.g. spliceosomal RNAs, miRNAs and snoRNAs
\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013};
these categories have been used to classify our results.

@@

%SnoRNA annotations overlapping miRNAs annotations where manually
%inspected \ppg{and?...}.

Chicken snoRNA and miRNA annotations were validated using small
RNA-seq data comprising 27 samples from 14 diferent tissues. Since not
all RNAs are expressed at any given time we expect to verify only a
fraction of annotated RNAs using the expression data. Nevertheless 242
of 691 (35\%) miRNAs as well as 328 of 376 (87\%) snoRNAs of our
homology-based annotations were found to be supported by the RNA-seq
data.

@@

Erich Jarvis (Duke University), Guojie Zhang (BGI-Shenzhen \&
University of Copenhagen) and Tom Gilbert (University of Copenhagen)
for access to data and for invaluable feedback on the manuscript.

Magnus Alm Rosenblad (Univ. of Gothenburg) for useful discussions.

We thank Fiona McCarthy (Mississippi State University) and Carl
Schmidt (University of Delaware) for providing the RNA-seq data.

%%%%%%%%%%%%%%%%%%%%%%
%% The Bibliography %%
%%
%% ---------------------------------
%% BioMedCentral bibtex .BST file will be used to
%% create a .BBL file which includes the BMC XML.
%% Note that the displayed Bibliography will not be
%% exactly as specified in the online Instructions for Authors

@@

\begin{figure}[ht]
  \includegraphics[width=0.75\textwidth]{figures/More-cand.pdf}
  \caption[]{Additional homologs of some sparsely represented RNA
    families were discovered using dedicated search strategies
    combined with highly sensitive settings, synteny information,
    lineage-specific CMs and subsequent manual
    inspection.}\label{fig:7}
\end{figure}

### 2013-11-05 paper/supplementary-material.tex
Updated manuscript, scripts, fixed chicken seqIds, ...
Making it look like an article, updated phylogenetic orderings.
Merged comments from Peter, Sarah, Sam & Mario.

Nearly all the microRNAs that are broadly conserved in fish,
amphibians and mammals are also conserved in the birds. Nevertheless,
there are obvious instances of microRNAs lost in all birds. For
example, mammalian, amphibian and fish genomes contain three loci of
clustered microRNAs from the mir-17 and mir-92 families. One of these
clusters (mir-93~25) was not found in turtles, crocodiles and
birds. In addition, we can confidently identify a further 3 microRNA
families that are present in mammals, and turtle and/or crocodile, but
not in any avian genome (mir-150, mir-208, mir-590). This suggests
that these sequences were lost in the last common ancestor of
archosaurs or birds. There are also a number of microRNAs that are
predicted to be present in turtles and/or crocodiles, and only a small
number of bird genomes. Indeed, there are many missing annotations,
species-specific and otherwise, that are not consistent with the
consensus phylogeny, and could be due to either incomplete genomes or
widespread microRNA loss.

@@

Bacterial families can be used to identify problematic sequences that
are likely to be the result of contamination from non-avian sources.

@@

\begin{figure}[ht]
  \includegraphics[width=0.45\textwidth]{figures/tRNA.pdf}
  \caption[]{Heatmaps showing the prescence/abscence and approximate
    copy number of {\bf tRNA} families.}\label{fig:5}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=0.45\textwidth]{figures/Y.pdf}
  \caption[]{...}\label{fig:6}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=0.45\textwidth]{figures/More-cand.pdf}
  \caption[]{...}\label{fig:7}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=0.45\textwidth]{figures/mir-17-cluster.pdf}
  \caption[]{...}\label{fig:8}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=0.45\textwidth]{figures/let-7-cluster.pdf}
  \caption[]{...}\label{fig:9}
\end{figure}

>>>>

In addition, we can confidently identify a further 3 microRNA
families that are present in mammals, and turtle and/or crocodile, but
not in any avian genome (mir-150, mir-208, mir-590). This suggests
that these sequences were lost in the last common ancestor of
archosaurs or birds. There are also a number of microRNAs that are
predicted to be present in turtles and/or crocodiles, and only a small
number of bird genomes. Indeed, there are many missing annotations,
species-specific and otherwise, that are not consistent with the
consensus phylogeny, and could be due to either incomplete genomes or
widespread microRNA loss.

@@

Bacterial families can be used to identify problematic sequences that
are likely to be the result of contamination from non-avian
sources. We identified a number of RNA families of bacterial origin in
the avian genomes. These have been reported and will be dealt with in
later updates to the avian genome sequences.

@@

\begin{figure}[ht]
  \includegraphics[width=0.85\textwidth]{figures/tRNA.pdf}
  \caption[]{Heatmaps showing the prescence/abscence and approximate
    copy number of {\bf tRNA} families.}\label{fig:5}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=0.85\textwidth]{figures/Y.pdf}
  \caption[]{...}\label{fig:6}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=0.85\textwidth]{figures/mir-17-cluster.pdf}
  \caption[]{...}\label{fig:8}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=0.85\textwidth]{figures/let-7-cluster.pdf}
  \caption[]{...}\label{fig:9}
\end{figure}
